---
title: "Medical Costs Analysis"
author: "Josué Ortiz & Olin Yoder"
date: "2023-03-29"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Libraries
```{r, message = FALSE}
library(tidyverse)
library(ggplot2)
library(stats)
library(lmtest)
library(gvlma)
library(regclass)
library(MASS)
library(leaps)
library(gbm)
library(caret)
library(knitr)
library(doBy)
library(rpart.plot)
library(magrittr)
library(caTools)
library(dplyr)
```

```{r pressure, include=FALSE}
costs <- read.csv('/Users/olinyoder/Desktop/Spring 2023/BSAN 440/Datasets/insurance.csv')
```

## Data Dictionary
https://www.kaggle.com/datasets/mirichoi0218/insurance
```{R}
data.desc <- data.frame(Variable = c("Age", "Sex","BMI","Children","Smoker","Region","Charges"),
                        Explanation = c("Age of primary beneficiary", 
                                        "Insurance contractor gender: Male or Female",
                                        "Body mass index",
                                        "Number of children covered by health insurance / Number of dependents",
                                        "Whether the individual smokes",
                                        "The beneficiary's US location: NE, SE, SW, NW",
                                        "Individual medical costs billed by health insurance"))
kable(data.desc)
```

## Data Highlights
##### Summary of variables
```{r}
kable(summary(costs))
```
##### NA values
```{r}
sum(is.na(costs))
```
There are no missing or null values in the data set.

##### Size of data set
```{r}
dim(costs)
```
There are 1338 observations and seven variables

### Medical Charges by Gender
```{r}
ggplot(costs, aes(x = sex, y = charges))+
  geom_boxplot()+
  xlab("Sex")+
  ylab("Charges")
```

### Is there a significant difference in the charges between males and females?

$$
{H_0}= \text{There is no difference in the average charge between males and females}\\
{H_a}= \text{The average charge for males is higher than females}
$$


```{r, echo = FALSE}
chargeSex <- summaryBy(charges  ~ sex, data=costs,na.rm=TRUE)
kable(chargeSex, col.names = c('Sex', 'Average Charge'))
```

```{r, message = FALSE}
ggplot(costs, aes(charges))+
  geom_histogram(color="white", fill="darkgray")+
  facet_grid(sex~.)+
  xlab("Charges")
```



Although there is not normality among charges by sex, since the data set is large enough, a t-test can be used.
```{r}
t.test(costs$charges ~ costs$sex, alternative = "greater")
```
Since the p-value, .9821, is greater than .05, we fail to reject that the average charge for males is higher than the average charge for females.

### Medical Charges by Region
```{r}
ggplot(costs, aes(x = region, y = charges))+
  geom_boxplot()+
  xlab("Region")+
  ylab("Charges")
```


### Is there a significant difference in the charges between regions?

$$
{H_0}= \text{There is no difference in the average charge between regions}\\
{H_a}= \text{The average charge for regions is different}
$$

```{r, message = FALSE}
ggplot(costs, aes(charges))+
  geom_histogram(color="white", fill="darkgray")+
  facet_grid(region~.)+
  xlab("Charges")
```

There is not normality among charges for each regions which could pose a potential problem for an anova test. Regardless, let's run an anova test without any transformation of charges.

```{r}
res.aov <- aov(charges ~ region, data = costs)
summary(res.aov)
```

Since the p-value, .0309, is less than .05, we reject that the average charges across regions are the same.

However, what if the normality among charges for each region were normally distributed?
```{r, message = FALSE}
ggplot(costs, aes(log10(charges)))+
  geom_histogram(color="white", fill="darkgray")+
  facet_grid(region~.)+
  xlab("Charges")
```

```{r}
res.aov2 <- aov(log10(charges) ~ region, data = costs)
summary(res.aov2)
```
Since the p-value, .241, is greater than .05, we fail to reject that the average charges across regions are the same. This contradicts our earlier conclusion, which could be a type 1 error (false-positive). 
Normally, anova tests handle non normal distributions in a manner where the false-positive rate (rejecting a true null) does not increase dramatically because of the skewness in the distribution; however, that is not the case for our data.


### How do other variables relate to charges?
#### Charges vs. Children
```{r}
ggplot(costs, aes(x = children, y = charges))+
  geom_point()+
  xlab("Children")+
  ylab("Charges")
```


There appears to be a slight negative relationship between a patient's charges and the number of children.

#### Charges vs. Age

```{r}
ggplot(costs, aes(x = age, y = charges))+
  geom_point()+
  xlab("Age")+
  ylab("Charges")
```


There is a positive, somewhat linear trend between age and charges. However, there are three distinctive lines suggesting that there is variables we need to take into consideration.


#### Charges vs. BMI

```{r}
ggplot(costs, aes(x = bmi, y = charges))+
  geom_point()+
  xlab("BMI")+
  ylab("Charges")
```


There is a moderate, positive, linear relationship between charges and BMI.

#### Charges vs. Smoking

```{r}
ggplot(costs, aes(x = smoker, y = charges))+
  geom_boxplot()+
  xlab("Smoker")+
  ylab("Charges")
```


Patients who smoke have a much higher median average charge than those who do not.

## Creating a Model to Predict a Patient's Charge



### Linear Regression

$$
{H_0}= \text{No variables are signficant in predicting a patient's charges}\\
{H_a}= \text{At least one variable is significant in predicitng a patient's charges}
$$

##### 1) Fit a saturated model
```{r}
saturated_lm <- lm(charges ~., costs)
summary(saturated_lm)
```
All variables except sex are significant in predicting a patient's charge.
Additionally, since the f-test returns a p-value of <2.2e-16, we reject the null hypothesis and conclude that at least one variable in significant in predicting charges.

```{r}
hist(costs$charges, main= "Distribution of Charges", xlab = "Charges")
shapiro.test(costs$charges)
```
Since charges is not normally distributed, a transformation may be necessary to meet the model assumptions.
### Box-Cox
```{r}
bc <- boxcox(charges ~ ., data=costs)
```

```{r}
lambda_bc2 <- bc$x[which.max(bc$y)]
lambda_bc2
```

Since lambda, .14, is close to 0, we can take a log transformation is adequate.

##### 2) Refit a saturated model, this time with a log transformation
```{r}
saturated_log_lm <- lm(log(charges) ~., costs)
summary(saturated_log_lm)
```
This time, all variables are significant. However, oftentimes a simpler model is better.

##### 3) Feature Selection
```{r}
# variance inflation factor - test for multicollinearity across predictor variables
VIF(saturated_lm)
```
Since all variables have a GVIF^(1/(2*Df)) near 1, there should not be an issue with multicollinearity. In that case, we will simply use forward stepwise selection instead of ridge, lasso, or elastic-net.

```{r}
forward <- regsubsets(log(charges) ~., costs, nvmax = 6, method = 'forward')
forward_summary <- summary(forward)

#using r^2
plot(forward_summary$adjr2, xlab="Number of Variables",
ylab="Adj r^2", type = "b", pch=17,col="black")

which.max(forward_summary$adjr2)
```
According to adj. r^2, the best model contains all six variables, which is the saturated log model from above. However, from the plot, we can see that a model with two variables performs nearly as well as a model with six variables. 

```{r}
coef(forward, 2)
```

##### 4) Create Model
```{r}
new_model <- lm(log(charges) ~ age + smoker, costs)
summary(new_model)
```

$$
ln(charges) = 7.38 + .036(age) + 1.547(smokerYes)
$$

##### 5) Model Diagnostics

```{r}
gvlma(new_model)
```

```{r}
par(mfrow=c(2,3))

plot(new_model)
hist(resid(new_model), main = "Distrbution of Residuals", xlab = "Residuals")
hist(stdres(new_model), breaks = 20, xlim = c(-5,5), main = "Distribution of Standardized Residuals", xlab = "Standardized Residuals")
abline(v = -3)
abline(v = 3)
```

Since the data does not follow the assumptions for a linear model well, we will instead take a tree-based approach.

### Decision Tree
To keep a constant response variable, we will take the log of charges again.
```{r}
sample_data = sample.split(costs, SplitRatio = 0.7)
train_data <- subset(costs, sample_data == TRUE)
test_data <- subset(costs, sample_data == FALSE)

dtmodel <- rpart(log(charges) ~ ., train_data)
rpart.plot(dtmodel)
```

### Boosted Model
```{r}
costs[sapply(costs, is.character)] <- lapply(costs[sapply(costs, is.character)], 
                                             as.factor)
set.seed(3456)
trainIndex <- createDataPartition(costs$charges, p = .7, 
                                  list = FALSE, 
                                  times = 1)
costsTrain <- costs[trainIndex,]
costsTest  <- costs[-trainIndex,]
```

```{r, warning = FALSE, message = FALSE}
costsGBM =gbm(log(charges)~.,data=costsTrain, distribution="gaussian",
                     n.trees=9000, interaction.depth=1, shrinkage = .001)
costs.best.iter <- gbm.perf(costsGBM, method = "OOB")
print(costs.best.iter)
```
The number of trees is set to 9000, although that many is not necessary.

```{r}
summary(costsGBM, plotit = FALSE)
```
Whether a person is a smoker and their age play the largest role in determining their charges.

### Which model is better? 
##### Boosted
```{r, message=FALSE}
costsPre = predict(costsGBM,newdata = costsTest)
costsPreRMSE <- sqrt(mean((costsPre - log(costsTest$charges))^2))
costsPreRMSE
```

##### Regression Tree
```{r}
train <- trainControl(method = "cv", number = 10)

dtRMSE = train(
   log(charges) ~ ., 
   data = costs, 
   method = "rpart",
   tuneGrid = expand.grid(cp = seq(from = 0, to = 0.1, by = 0.01)),
   metric = "RMSE",
   trControl = train
)

dtRMSE
```
At cp = 0, RMSE is the lowest.

#### Linear Regression
```{r}
cv <- train(log(charges) ~ age + smoker, data = costs, method = 
               "lm", trControl = train, na.action = na.omit)
cv
```

Although all the models have a similar RMSE, .39 - .46, the linear regression model is misleading since the data is not well explained linearly. Between the regression tree and boosted model, the regression tree, even without pruning, has a surprisingly lower RMSE which suggests it is the better model. However, in reality, it is likely overfit.

